\documentclass[12pt]{book}
\raggedbottom
\usepackage[top=1in,left=1in,bottom=1in,right=1in,headsep=0.25in]{geometry}	
\usepackage{amssymb,amsmath,amsthm,amsfonts}
\usepackage{chapterfolder,docmute,setspace}
\usepackage{cancel,multicol,tikz,verbatim,framed,polynom,enumitem,tikzpagenodes}
\usepackage[colorlinks, hyperindex, plainpages=false, linkcolor=blue, urlcolor=blue, pdfpagelabels]{hyperref}
\usepackage[type={CC},modifier={by-sa},version={4.0},]{doclicense}

\theoremstyle{definition}
\newtheorem{example}{Example}
\newcommand{\Desmos}{\href{https://www.desmos.com/}{Desmos}}
\setlength{\parindent}{0in}
\setlist{itemsep=0in}
\setlength{\parskip}{0.1in}
\setcounter{secnumdepth}{0}
\input{lesson_order}

\newcommand{\tmmathbf}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\tmop}[1]{\ensuremath{\operatorname{#1}}}

\begin{document}
\section{Graphing Summary (L\arabic{lesson_polynomials_graphing_summary})}
{\bf Objective: Graph a polynomial function in its entirety.}\par
At this point, we have addressed all key features of polynomials individually.  This section pulls each of these aspects together, for a detailed analysis of a polynomial, culminating in a complete sketch of its graph.  Along the way, we will need to address each of the following aspects for our polynomial $f(x)=a_nx^n+a_{n-1}x^{n-1}+\ldots+a_1x+a_0$.  It is important to note that there is no universally accepted order to this checklist.
\begin{itemize}
	\item Find the $y-$intercept of the graph of $f,$ $(0,f(0))=(0,a_0)$.
	\item Use the degree $n$ and leading coefficient $a_n$ to determine the end behavior of the graph of $f$.
	\item Identify a complete factorization of $f,$ and use it to find any $x-$intercepts of the graph of $f$.  Using multiplicities, classify each $x-$intercept as a crossover or turnaround (``bounce'') point.
	\item Using the $x-$intercepts, construct a sign diagram for $f$.
\end{itemize}
In each polynomial we encounter, we will carefully examine the function, making sure not to omit any of the checklist items above and to compare each item to those that precede it along the way for accuracy.  Although the process will take some time, if we are thorough, our end result should be a complete, accurate sketch of the given polynomial.\par
\begin{example}~~Sketch a complete graph of $f(x)=14x^4-17x^3-6x^2+7x+2$.\par
We will start with the $y-$intercept, which is $(0,2)$.\par
Next, we see that $f$ has even degree and positive leading coefficient.  So, the tails of the graph of $f$ both point upwards.  In other words, as $x\rightarrow\pm\infty, f(x)\rightarrow\infty$.\par
Since $f$ is degree-4, contains more than four terms, and is not of quadratic type, we will apply the Rational Root Theorem.  In this case, our set of possible rational roots is
$$\left\{\pm 1, \pm 2, \pm \frac{1}{2}, \pm \frac{1}{7}, \pm \frac{1}{14}, \pm \frac{2}{7}\right\}$$
Fortunately, we see that $f(1)=14-17-6+7+2=0$.  So, $x-1$ is a factor of $f$.  Dividing, we get:
\begin{multicols}{2}
\polylongdiv{14x^4-17x^3-6x^2+7x+2}{x-1}

\columnbreak

\polyhornerscheme[x=1,showbase=top,resultstyle=\bf]{14x^4-17x^3-6x^2+7x+2}
\end{multicols}
So, $f(x)=(x-1)(14x^3-3x^2-9x-2).$  Applying the Rational Root Theorem a second time, we can see that $x=1$ is also a root of the cubic factor of $f,$ since $14-3-9-2=0$.  Again, we can divide to factor $f$ further.
\begin{multicols}{2}
\polylongdiv{14x^3-3x^2-9x-2}{x-1}

\columnbreak

\polyhornerscheme[x=1,showbase=top,resultstyle=\bf]{14x^3-3x^2-9x-2}
\end{multicols}
So, $f(x)=(x-1)^2(14x^2+11x+2)$.  Factoring the remaining quadratic, we have
$$f(x)=(x-1)^2(7x+2)(2x+1),$$
with accompanying set of roots $\{1,-\frac{1}{2},-\frac{2}{7}\}$.\par
Using multiplicities, we conclude that the $x-$intercept $(1,0)$ is a turnaround point, and the intercepts $\left(-\frac{1}{2},0\right)$ and $\left(-\frac{2}{7},0\right)$ are crossover points.\par
Though not necessary for graphing, a sign diagram confirms our end and local behavior findings.
\begin{center}
\begin{tikzpicture}[xscale=1,yscale=1]
	\draw [<->](-6.25,0) -- coordinate (x axis mid) (4.25,0) node[below right] {$x$};
	\draw [-](-4,1) -- coordinate (y axis mid) (-4,-0.25) node[below] {$-\frac{1}{2}$};
	\draw [-](-1,1) -- coordinate (y axis mid) (-1,-0.25) node[below] {$-\frac{2}{7}$};
	\draw [-](2,1) -- coordinate (y axis mid) (2,-0.25) node[below] {$1$};
	\draw (-5.5,-1) node {$x=-1$};
	\draw (-2.5,-1) node {$x=-\frac{3}{7}$};
	\draw (0.5,-1) node {$x=0$};
	\draw (3.5,-1) node {$x=2$};
	\draw (-5.5,0.5) node {$+$};
	\draw (-2.5,0.5) node {$-$};
	\draw (0.5,0.5) node {$+$};
	\draw (3.5,0.5) node {$+$};
\end{tikzpicture}
\end{center}
Putting all of this information together results in the following graph.
\begin{center}
\begin{tikzpicture}[xscale=1.75,yscale=1]
	\draw [<->](-2.25,0) -- coordinate (x axis mid) (2.25,0) node[below right] {$x$};
	\draw [<->](0,-1.75) -- coordinate (y axis mid) (0,5.25) node[above right] {$y$};
	\draw [<->] plot [domain=-0.736:1.33, samples=100] (\x,{(2*\x+1)*(7*\x+2)*(\x-1)*(\x-1)});
	\foreach \x in {0.5,1,1.5,2}
		\draw (\x,1pt) -- (\x,-1pt)	node[anchor=north] {\scriptsize \x};
	\foreach \x in {-2,-1.5,-1,-0.5}
		\draw (\x,1pt) -- (\x,-1pt)	node[anchor=north] {\scriptsize \x};
	\foreach \y in {1,...,5}
		\draw (1pt,\y) -- (-1pt,\y)	node[anchor=east] {\scriptsize \y}; 
	\foreach \y in {-1}
		\draw (1pt,\y) -- (-1pt,\y)	node[anchor=west] {\scriptsize \y}; 
\end{tikzpicture}
\end{center}
\end{example}
\end{document}